A Practical Optional Type System for Clojure

### 2012-10-22 dissertation/design-choices.tex
Example for infer
All of Rowan's corrections

The majority of the design and implementation of Typed Clojure is based on Typed Racket.

@@

\subsection{Interaction with null}
% null is explicitly expressible in my type system
% In Java, null is subtype of all reference types, any reference type must be tested to prevent NPE
%

@@

\section{Multimethods}

% INTRODUCTION TO CLOJURE MULTIMETHODS

%A multimethod in Clojure differs from most other
%abstractions of the same name. Clojure multimethods
%dispatch on the result of an arbitrary function
%rather than the types or structure of its arguments.
%
%\begin{lstlisting}[caption=Clojure Multimethods, label=lst:mm1]
%(defmulti ToString class)
%
%(defmethod ToString java.lang.Integer
%  [a]
%  (str "Integer: " a))
%
%(defmethod ToString java.lang.Number
%  [a]
%  (str "Number: " a))
%\end{lstlisting}
%
%Even with this extra flexibility, multimethods are often used to dispatch on the type of its arguments.
%Listing \ref{lst:mm1} illustrates a common usage of multimethods: dispatching on the class of the first argument.
%We use \lstinline|defmulti| to define a new multimethod.
%The function \lstinline|class| is used as the dispatch function, a one-argument function returning the class of its argument
%as an instance of \lstinline|java.lang.Class|.
%
%A multimethod invocation redirects control flow to exactly one of its \emph{methods} with unresolved ambiguities resulting in
%a runtime error. The method is selected by first comparing the result of the dispatch function to
%the dispatch values defined by each method. If one or more methods match, multimethods provide
%the ability to prefer one method over another.

@@

Similar to Typed Racket, type inference is based on Local Type Inference
by Pierce and Turner.

@@

Typed Clojure includes support for F-bounds on type variables, as an extension
to Local Type Inference.

@@

>>>>

The majority of the design and implementation of Typed Clojure is based on Typed Racket.
This is a deliberate choice: the purpose of this work is to produce
a practical tool, and there is already good evidence that Typed Racket
is practical and makes good tradeoffs and design choices.
Our differences are mostly due to the differences between Clojure and Racket.

@@

\subsection{Interaction with null}

It is a goal of Typed Clojure to statically prevent many Java \emph{NullPointerException}
exceptions.
\emph{null} is explicitly expressible in Typed Clojure as the type \listline|nil|.
This enables more expressive types that specify where \emph{null} is and is not allowed.
Section \ref{sec:javainterop} gives an example of how this could be useful 
when using Java interoperability and gives details on the different ways Typed Clojure.

%===> Occurrence typing~\cite{TF10} also plays a crucial role in Typed Clojure's

@@

@@

Similar to Typed Racket, type inference is based on Local Type Inference
by Pierce and Turner~\cite{PT00}.

@@

Typed Clojure includes support for F-bounds on type variables\cite{CCHOM89}, as an extension
to Local Type Inference.

@@

% what about get? Core Type Rules

### 2012-10-22 dissertation/future-work.tex
All of Rowan's corrections

\section{Multimethods}
% Problem - typing isa?
% - probably impossible to type higher order usages of isa?.
% Dispatch function must infer latent filters for multi-arity functions. (Typed Racket only infers single arg I think)
% Statically proving at least one relevant dispatch method is present.
% Warn potential method conflicts.
% Suppress warnings, similar to prefer-method syntax

\section{Records}
% A composition of existing features: Datatypes and Heterogeneous Maps

@@

>>>>

\section{Contracts and Blame}
\label{sec:contractsblame}

\section{Multimethods}
\label{sec:mmfuture}

\begin{verbatim}
Multimethods in Clojure are an idiomatic and often used feature.
It would be essential for an optional static type system for
Clojure to support
There are at least two hurdles to supporting multimethods.
\end{verbatim}

% Problem - typing isa?
% - probably impossible to type higher order usages of isa?.
% Dispatch function must infer latent filters for multi-arity functions. (Typed Racket only infers single arg I think)
% Statically proving at least one relevant dispatch method is present.
% Warn potential method conflicts.
% Suppress warnings, similar to prefer-method syntax

\section{Records}

A Clojure record is a composite of datatypes and maps; it is a datatype
with fields that can be treated like a map (records support map operations like \lstinline|get| (lookup),
\lstinline|assoc| (add entries), and \lstinline|dissoc| (remove entries)).
One interesting property from the perspective of static typing is that if
a entry corresponding to a field of the underlying datatype is removed
with \lstinline|dissoc|, then the resulting type is a plain map: it loses
its record type.
Correspondingly, a record keeps its type if any other entry is removed.

Records are used frequently in Clojure code, so it is desirable
to support them to some degree. Future work is planned to investigate solutions
to satisfactorily capture their subtle semantics in a practical way.

@@

\begin{verbatim}
Other Clojure implementations
\end{verbatim}

\subsection{Java Generics}

\begin{verbatim}
Currently, Typed Clojure does not interact with Java generics,
which provide
instead 
It may be desirable to 
\end{verbatim}

### 2012-10-22 dissertation/intro.tex
All of Rowan's corrections

It is common in Clojure, like other dynamically-typed languages, to
encode implicit type invariants in conditional tests.
In listing \ref{lst:numvec2} conditional tests are used
to refine the type of the bindings \lstinline|a| and \lstinline|b|.
Occurrence typing is a technique useful for capturing these kinds
of type invariants (occurrence typing is discussed in further detail in section \ref{sec:OccurrenceTyping}).

@@

\subsection{Java Interoperability}

Typed Clojure currently provides safe interoperability
\footnote{Compatibility with Generic Java is planned future work}.

The key to safe Java interoperability is the treatment of Java's \emph{null}.
\emph{null} is a subtype to all reference types
represented in Clojure by \lstinline|nil|. Unlike Java's type system
Clojure explicitly separates \emph{null} and reference types, allowing
Typed Clojure to express
\emph{nullable}
\footnote{A type is \emph{nullable} if it may also be an instance of \emph{null},
which is expressed in Typed Clojure by creating a union of the reference type and \lstinline|nil|.}
positions.

@@

This example shows how Typed Clojure handles \emph{null} while creating and
using an instance of \emph{java.io.File}\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.

@@

Next, we see how Typed Clojure's default behaviour treats method return positions as nullable.
The \emph{java.io.File} instance method \emph{getParent}
is equivalent to \lstinline|[-> (U nil String]| in Typed Clojure. This happens to be
a valid approximation of the method as \emph{getParent} returns \emph{null} 
``if the pathname does not name a parent directory''\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
On the other hand, the instance method \emph{getName} always returns an
instance of \emph{java.lang.String}, so we set the return position of
\emph{getName} to non-nil with \lstinline|typed.core/non-nil-return|
\footnote{The second parameter to \lstinline|non-nil-return| specifies which arities to assume non-nil 
returns, accepting either a set of parameter counts of the relevant arities, or \lstinline|:all|
to override all arities of that method.}.

>>>>

It is common in Clojure, like other dynamically-typed languages, to
encode implicit type invariants in conditional tests.
In listing \ref{lst:numvec2} conditional tests are used
to refine the type of the bindings \lstinline|a| and \lstinline|b|.
Occurrence typing\cite{TF10} is a technique useful for capturing these kinds
of type invariants (occurrence typing is discussed in further detail in section \ref{sec:OccurrenceTyping}).

@@

\subsection{Java Interoperability}
\label{sec:javainterop}

Typed Clojure supports integration with Java's type system during
interoperability from Clojure. This offers many of the same advantages
as using Java's type system from Java, for example it is a type
error to pass arguments of the wrong type to methods.
Typed Clojure also makes different decisions to Java's type system,
particularly in the treatment of Java's \emph{null}.

From the perspective of static types in Java, \emph{null} is included in all reference types.
\emph{null} is represented in Clojure by the value \lstinline|nil|. Unlike Java's type system
Typed Clojure explicitly separates \emph{null} and reference types giving Typed Clojure 
more accurate types: we can directly express \emph{nullable} and \emph{not nullable} types.
(A type is \emph{nullable} if it may also be \emph{null},
which is expressed in Typed Clojure by creating a union of the type and \lstinline|nil|.
A type is \emph{not nullable} if it doesn't include \emph{null}).

@@

This example shows how Typed Clojure handles \emph{null} while creating and
using an instance of \emph{java.io.File}.

@@

Next, we see how Typed Clojure's default behaviour treats method return positions as nullable.
The \emph{java.io.File} instance method \emph{getParent}
is equivalent to \lstinline|[-> (U nil String]| in Typed Clojure. This happens to be
a valid approximation of the method as \emph{getParent} returns \emph{null} 
``if the pathname does not name a parent directory''\footnote{See Javadoc for \emph{java.io.File}: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
On the other hand, the instance method \emph{getName} always returns an
instance of \emph{java.lang.String}, so we set the return position of
\emph{getName} to non-nil with \lstinline|typed.core/non-nil-return|
(the second parameter to \lstinline|non-nil-return| specifies which arities to assume non-nil 
returns, accepting either a set of parameter counts of the relevant arities, or \lstinline|:all|
to override all arities of that method).

Such annotations are in a sense assumptions; should they turn out to be wrong, Typed Clojure
will infer incorrect types. Typed Racket generates runtime contracts based on static types to ensure
that type errors are caught at runtime, and the original source of the error is blamed~\cite{WF09}.
A similar system for verifying that method annotations like \lstinline|non-nil-return|
actually describe correct types is planned as future work
(see section \ref{sec:contractsblame}).

### 2012-10-22 dissertation/lit-review.tex
All of Rowan's corrections

In this chapter we fit the design of Typed Clojure with literature from related fields.
Forward-references are given to link the discussed literature with this project
where there is some relevance or influence.

@@

% cite: Dynamic Typing in Polymorphic languages
% - coming from the other direction
% - extending static languages to be more dynamic

@@

There is a history of designing static type systems for untyped languages.

@@

Type systems for Scheme have since moved away from soft typing
to to other approaches like gradual typing.
Alas, the best reference for commentary on this transition 
appears to be slides from a talk by Felleisen~\cite{Fell09},
a leader in this area of research for over 20 years.
Felleisen says that while Soft Scheme discovered type problems, 
it suffers from incomprehensible
errors that required PhD-level expertise to decipher.

@@

In typed languages like Haskell, functions are as simple as possible, taking a single argument.
A function with multiple arguments is represented by chaining several single-argument function
together, or by using lists, or using tuples. This style is characterised by direct syntactic function currying, where applying
a function to less than its maximum number of arguments results in another function
that takes the remaining arguments.

@@

\begin{lstlisting}[caption=An application of the non-uniform variable-arity function \lstinline|map|, label=lst:map]
(map + [1 2] [2.1 3.2]) 
;=> (3.1 5.2)
\end{lstlisting}\footnote{Line comments in Clojure begin with \lstinline|;| and comments to the end of the line. We use \lstinline|;=>| to mean \emph{evaluates to}.}

@@

Occurrence typing~\cite{TF08,TF10} extends the type 
system with a \emph{proposition environment} that represents 
the type refinements inferred down a particular path.
The 2010 paper~\cite{TF10}
reformulated occurrence typing, improving the original 2008 paper
~\cite{TF08}
after revealing that ``three years of practical experience has revealed
serious shortcoming of our type system.''~\cite{TF10}
The new implementation utilizes a simple proof system to solve
propositional logic statements, in terms of type predicates and selectors.

@@

An exciting application of occurrence typing as yet unexplored is facilitating null-safe interop with Java.
By declaring \lstinline|nil| (Clojure's value of Java's \lstinline|null|) to \emph{not} be a subtype of reference types,
we can begin to statically disallow potential inconsistent usages of \lstinline|nil| as part of the type system.

@@

Listing \ref{lst:nil} infers from the Java signature \lstinline|Object getObject()| that
that \lstinline|a| is of type \lstinline|(U nil Object)|. This is equivalent to Java's
\lstinline|Object| static type, as \lstinline|null| is a subtype of all reference types. By surrounding
the call \lstinline|(expects-non-nil a)| with \lstinline|(when a ...)|, we guarantee that
\lstinline|a| is non-nil when passed to \lstinline|expects-non-nil|. Occurrence typing infers
this by observing \lstinline|nil| is a false value in Clojure, therefore \lstinline|a| cannot
be \lstinline|nil| the body of the \lstinline|when|, refining \lstinline|a|'s type to \lstinline|Object|
from \lstinline|(Union nil Object)|.

Occurrence typing is a relatively simple, time worn technique used successfully 
in Typed Racket. Clojure is similar enough to Racket for occurrence typing to work
without issues, and has good potential to enable null-safe Java interop.

@@

Many advanced type systems provide support for \emph{higher kinds},
which are ``another level up'' from types.
For example, a type \lstinline|Number| is distinguished from a
\emph{type constructor} \lstinline|Monad| (see section \ref{sec:monads}), which is a type level \emph{function}
(similar to the difference between values and functions on the value level, but on the type level).

>>>>

In this chapter we fit the design of Typed Clojure with literature from related fields.
Forward-references are often given to link the discussed literature with this later parts
of this dissertation
where there is some relevance or influence.

@@

% cite: Dynamic Typing in Polymorphic languages (1991?)
% - coming from the other direction
% - extending static languages to be more dynamic

@@

This section briefly reviews work on designing static type systems for untyped languages.

@@

% Rowan
% You've missed some of the steps. Most importantly the move to program analysis
% techniques like Set Based Analysis (SBA) - which suffered from modularity issues
% due to the lack of declared intended types.
% Maybe quote Felleisen - "problem: Mr Spidey and friends infer brittle and large
% types; errors remain difficult to explain and fix"

Type systems for Scheme have since moved away from soft typing
to to other approaches like gradual typing.
Alas, the only reasonably complete reference for commentary on this transition 
appears to be slides from a talk by Felleisen~\cite{Fell09},
a leader in this area of research for over 20 years.
Felleisen says that while Soft Scheme discovered type problems, 
it suffers from incomprehensible
error reporting that required PhD-level expertise to decipher.

@@

In typed languages like Haskell, functions are as simple as possible, taking a single argument.
A function with multiple arguments is represented by chaining several single-argument function
together, or by using lists, or using tuples. This first style is characterised by direct syntactic function currying, where applying
a function to less than its maximum number of arguments results in another function
that takes the remaining arguments.
Using tuples instead requires that the same number of arguments be supplied,
while using lists allows any number, but requires that all arguments have the same type.

@@

\begin{lstlisting}[caption=An application of the non-uniform variable-arity function \lstinline|map|, label=lst:map]
(map + [1 2] [2.1 3.2]) 
;=> (3.1 5.2)
\end{lstlisting}

(Line comments in Clojure begin with \lstinline|;| and comments to the end of the line. We use \lstinline|;=>| to mean \emph{evaluates to}).

@@

Occurrence typing~\cite{TF08,TF10} extends the type 
system with a \emph{proposition environment} that represents 
the information on the types of bindings down conditional branches.
These propositions are then used to update the types associated
with bindings in the \emph{type environment} down branches
so binding occurrences are given different types 
depending on the branches they appear in, and the conditionals
that lead to that branch.
(See section \ref{sec:occurenceimpl} for details on how Typed Clojure uses occurrence typing).

@@

An exciting application of occurrence typing as yet unexplored is facilitating null-safe interoperability with Java.
By declaring \lstinline|nil| (Clojure's value of Java's \lstinline|null|) to \emph{not} be a subtype of reference types,
we can begin to statically disallow potentially inconsistent usages of \lstinline|nil| as part of the type system.

@@

Listing \ref{lst:nil} infers from the Java signature \lstinline|Object getObject()| that
that \lstinline|a| is of type \lstinline|(U nil Object)|. This is equivalent to Java's
\lstinline|Object| static type, as \lstinline|null| is a member of all reference types. By surrounding
the call \lstinline|(expects-non-nil a)| with \lstinline|(when a ...)|, we guarantee that
\lstinline|a| is non-nil when passed to \lstinline|expects-non-nil|. Occurrence typing infers
this by observing \lstinline|nil| is a false value in Clojure, therefore \lstinline|a| cannot
be \lstinline|nil| the body of the \lstinline|when|, refining \lstinline|a|'s type to \lstinline|Object|
from \lstinline|(U nil Object)|.

Occurrence typing is a relatively simple technique used successfully 
in Typed Racket. Clojure is similar enough to Racket for occurrence typing to work
without issues, and has good potential to help programmers avoid using
\lstinline|nil| incorrectly

@@

Many advanced type systems provide support for \emph{higher kinds},
which are ``another level up'' from types.
For example, a type \lstinline|Number| is distinguished from a
\emph{type constructor} \lstinline|Monad| (see section \ref{sec:monads}), which is a type level function
(similar to the difference between values and functions on the value level, but on the type level).
